An Analysis of the Usage of the 2.4 GHz and 5 GHz Radio Bands in Antwerp

### 2017-05-26 advnetwpaper/paper_no_ieeetrans.tex
A few minor changes to spelling and wording.

How bad is it, in general? 
Are there patterns, e.g. daily recurrence? 
Advanced: how can you ensure data quality, i.e., how to validate the results?

@@

We perform one scan per minute, alternating between 2.4 GHz and 5 GHz. As each scan takes at least 25 seconds to complete, we chose not to lower this interval any further. This generates around 860MB of raw data per node per day. Even when only considering the maximum signal strength per sample, we still need some way to condense this to a more digestible format. We experimented with some smoothing algorithms and had the best results with the Savitzky-Golay filter\footnote{\url{https://en.wikipedia.org/wiki/Savitzky\%E2\%80\%93Golay_filter}}. A detailed analysis of the filter was considered outside of the scope of this document, but we note that it is a digital filter intended for equally spaced data points that smooths data through convolution. We always used polynomials of degree 3. The only other parameter is the window size. The higher the window size, the smoother the final result. Figure \ref{savitzky_0} shows the resulting smoothings for 2 different window sizes. The smaller window size reveals some details such as the drop in activity between noon and 1PM (lunch time), while the larger window size is detailed enough to show that daytime is more active than nighttime. This filter performs well mainly for the center of the data set. The edges of the smoothing are clearly inaccurate, especially the sudden rise at the end of the day. This is a known issue with this filter. Luckily we usually also have the data of the preceding and following day. By starting the smoothing earlier and ending it later, we obtain a better smoothing. Figure \ref{savitzky_1} was generated using the data starting at 11PM the day before up to 1AM the day after. In figure \ref{savitzky_4} we increased this margin to 4 hours at each end. While the 1 extra hour does little to improve the smoothing and even leads to a worse result in one case, the effect disappears completely with the 4 additional hours. We use this 4 hour buffer for all smoothings. Furthermore we decided to use the larger window size. The smoother graphs make it easier on the eye to compare multiple smoothings on one graph. Additionally effects such as the drop during lunchtime were often not seen with even tighter window sizes.\\
\newpage %Required to fix wrapfigure formatting
\begin{wrapfigure}{r}{0.35\textwidth}
\centering
\captionsetup{justification=centering}
\includegraphics[width=0.25\textwidth]{legend.png}
\caption{Legend:\\ days of the week}
\label{legend}
\end{wrapfigure}
Using an adapted version of Simon W{\"u}nderlich's FFT$\_$eval code, Python's Matplotlib library, the mathematical formulas mentioned above and a Savitzky-Golay filter with window size 361 we graphically represent our data points per node. For one week (Monday 15 May - Sunday 21 May) we plot each day's smoothing, for easy comparison between the days. For the 2.4 GHz band we perform this separately for the non-overlapping channels 1, 6 and 11. Each color in the plot represents a day of the week, using the legend in figure \ref{legend}. All other plots use the same legend, although days may be missing.

@@

\subsection{5 GHz band}
\subsubsection{Recurring patterns between days}
\begin{figure}
\begin{minipage}{0.47\textwidth}
    \centering
	\includegraphics[width=\textwidth]{images/2_4_GHz/node1_2017-05-17_chan1_image}
    \caption{Node 1, ch. 1, 13/05 - 17/05} \label{node1-1}
\end{minipage}\hfill
\begin{minipage}{0.47\textwidth}
    \centering
	\includegraphics[width=\textwidth]{images/2_4_GHz/node1_2017-05-17_chan6_image}
    \caption{Node 1, ch. 6, 13/05 - 17/05} \label{node1-6}
\end{minipage}\hfill
\begin{minipage}{0.47\textwidth}
    \centering
	\includegraphics[width=\textwidth]{images/2_4_GHz/node1_2017-05-17_chan11_image}
    \caption{Node 1, ch. 11, 13/05 - 17/05} \label{node1-11}
\end{minipage}\hfill
\begin{minipage}{0.47\textwidth}
\centering
\includegraphics[width=\textwidth]{images/5_GHz/cot-node3-student_2017-05-22_chan36_image.pdf}
\caption{Node 3, ch. 56, 15/05 - 22/05}
\label{cot3_5ghz_pattern_chan36}
\end{minipage}\hfill
\end{figure}
The data gathered during the 5 GHz measurements proved to be less interesting than the data we gathered in the 2.4 GHz measurements. Most of the channels we measured turned out to be silent or contain very little signal. This is partly to be expected. Many common communication devices only use the 2.4 GHz, and many routers in use do not yet support the 5 GHz band. Furthermore, any communication that does occur in this band can be more spread out as the band is a lot wider than the 2.4 GHz band. For the most part, the channels that did show traffic did not show any significant patterns or recurring phenomena. In most cases the received signal strength was very stable on all channels. We did notice some remarkable behaviour though. For example, node 3 measured a lower signal strength for some time around noon on both May 17\textsuperscript{th} and May 22\textsuperscript{nd} on channel 56, in figure \ref{cot3_5ghz_pattern_chan36}. The reason for this however is not as clear as with the daily peaks experienced in the 2.4 GHz It might be that the decrease occurs every couple of days, for example due to a router switching channels temporarily. It could also simply be an anomaly.

For this node, channel 56 was the only somewhat active channel. All other channels saw barely any traffic in the 5 GHz band. \\\\
It is worth noting that, apart from channel 36, there is some difference in the stability of the received signal between the outdoor node and the indoor nodes. The outdoor node has at least some activity on every channel whereas most of the indoors nodes show the bare minimum of activity for most channels. An example of the outdoors node's activity on channel 56 versus an indoors node on that same channel is shown in figures \ref{node1_5ghz_traffic_chan56} and \ref{node9_5ghz_traffic_chan56}. There could be some outside source of slight intermittent interference. As we only had access to one outdoors node we cannot determine whether this generally occurs with outside nodes, or is specific to this one.

@@

There were some other interesting phenomena we observed in the 5 GHz channels. For example, node 6 shows a sudden drop in activity on both channels 56 and 64, coinciding with an increase in traffic on channels 112 and 120, as seen in figures \ref{node6_5ghz_traffic_chan56} up to \ref{node6_5ghz_traffic_chan120}. Looking at a random spectral scan from before and after the shift, we indeed see the same behaviour. There was clear activity in channels 56 and 64 before the shift, seen in figure \ref{cot6_50_early}.
\begin{figure}[ht]
\begin{minipage}{0.47\textwidth}
	\centering
	\includegraphics[width=\textwidth]{images/5_GHz/cot-node6-student_2017-05-20_chan56_image.pdf}
	\caption{Node 6, ch. 56, 14/05 - 20/05}
	\label{node6_5ghz_traffic_chan56}
\end{minipage}\hfill
\begin{minipage}{0.47\textwidth}
	\centering
	\includegraphics[width=\textwidth]{images/5_GHz/cot-node6-student_2017-05-20_chan64_image.pdf}
	\caption{Node 6, ch. 64, 14/05 - 20/05}
	\label{node6_5ghz_traffic_chan64}
\end{minipage}\hfill
\begin{minipage}{0.47\textwidth}
    \centering
    \includegraphics[width=\textwidth]{images/5_GHz/cot-node6-student_2017-05-20_chan112_image.pdf}
    \caption{Node 6, ch. 112, 14/05 - 20/05}
    \label{node6_5ghz_traffic_chan112}
\end{minipage}\hfill
\begin{minipage}{0.47\textwidth}
    \centering
    \includegraphics[width=\textwidth]{images/5_GHz/cot-node6-student_2017-05-20_chan120_image.pdf}
    \caption{Node 6, ch. 120, 14/05 - 20/05}
    \label{node6_5ghz_traffic_chan120}
\end{minipage}\hfill
\end{figure}
 While there is some very slight activity in channel 60, so few samples contained any trace of it that it was filtered out. Later on, figure \ref{cot6_50_late} shows that this traffic signature has moved to channels 112 and 120, now with the channel in between showing no traffic at all. We are not sure what device exactly causes this, but it is clear that it was at some point switched to another channel. This shows that the activity in a channel can change at a moment's notice, and indicates that systematic or even constant monitoring of the spectrum is not without merit. We do note that the received high signal is still only -70dBm, meaning there is still room for communication. Furthermore the spectral scan samples show that the those channels are still silent for a large fraction of the time.

@@

\end{figure}
\subsection{Implications of high signal strength}
A high or unstable measured signal strength does not necessarily make a channel unusable for a City of Things node. First of all, we note that all received signal strengths previously reported was due to activity caused by external sources (i.e. not the node itself). When considering using the node for communication, all those signals are considered as noise. As long as this noise is lower than the signal strength used for the node's communication, there is no issue. While a constant noise of -20dBm will make communication very difficult, noise of -70dBm, -60dBm or even -50dBm should be no problem for close-range communication, as the node and its peer could probably generate a stronger signal. Furthermore it is important to note that in many cases the measured high signal strength was not constant. As it was probably caused by a communication device (likely Wi-Fi), the signal source remains silent for a significant fraction of the time. Furthermore such sources of potential interference probably implement some sort of listen-before-talk, further facilitating communication with the node in the presence of such sources. It is however difficult to predict exactly how well communication will work for a specific situation without actually trying it out.
\subsection{Validation}
While there is no official confirmation from the chip manufacturer that our interpretation of the spectral scan data is correct, we are confident that our interpretation gives at least a solid estimation of spectrum activity. First of all the experiment in section \ref{accuracy} shows that when we add a source of RF activity near the node, this is clearly seen in the spectral scan at exactly the right frequency. \\
Furthermore we discovered daily patterns where we expected them. The Middelheim offices are busy during the daytime on weekdays but fairly silent during the weekends and nights. While not shown in the plots above, holidays were just as quiet as weekends.
\section{Conclusions}
In this report we presented our findings on activity in the RF spectrum in the 2.4 GHz and 5 GHz bands in the city of Antwerp. One clear theme in our results is that the 5 GHz is overall quieter than the 2.4 GHz band. This is not surprising. The 5 GHz band is a lot wider than the 2.4 GHz band, and its signals do not carry as far. Furthermore, assuming that a lot of interference originated from Wi-Fi routers, one factor could be that many active routers and other devices do not yet support 5 GHz. In addition, technologies such as Bluetooth, which could also be creating signals, only affects 2.4 GHz.\\ \\
In some environments we notice a daily pattern where the signal strength is about 20dBm higher during the day compared to the night. As this was mainly observed on nodes in office spaces, this effect disappeared during the weekend. Other nodes show absolutely no difference between daytime and nighttime. Often this is when the spectrum is just very quiet. One interesting observation is that channels 6 and 11 were never busier than channel 1. \\
Furthermore, we found nothing indicating that City of Things network would experience considerable interference in the 5 GHz band, especially for the higher channel numbers. When operating on the 2.4 GHz band we observed more activity that could, in some places and on some channels, hinder communication with faraway hosts. However there was always at least one channel fairly silent.
\bibliographystyle{IEEEtran}
\bibliography{IEEEabrv,./bibliography}

>>>>

@@

We perform one scan per minute, alternating between 2.4 GHz and 5 GHz. As each scan takes at least 25 seconds to complete, we chose not to lower this interval any further. This generates around 860MB of raw data per node per day. Even when only considering the maximum signal strength per sample, we still need some way to condense this to a more digestible format. We experimented with some smoothing algorithms and had the best results with the Savitzky-Golay filter\footnote{\url{https://en.wikipedia.org/wiki/Savitzky\%E2\%80\%93Golay_filter}}. A detailed analysis of the filter was considered outside of the scope of this document, but we note that it is a digital filter intended for equally spaced data points that smooths data through convolution. We always used polynomials of degree 3. The only other parameter is the window size. The higher the window size, the smoother the final result. Figure \ref{savitzky_0} shows the resulting smoothings for 2 different window sizes. The smaller window size reveals some details such as the drop in activity between noon and 1PM (lunch time), while the larger window size is detailed enough to show that daytime is more active than nighttime. This filter performs well mainly for the center of the data set. The edges of the smoothing are clearly inaccurate, especially the sudden rise at the end of the day. This is a known issue with this filter. Luckily we usually also have the data of the preceding and following day. By starting the smoothing earlier and ending it later, we obtain a better smoothing. Figure \ref{savitzky_1} was generated using the data starting at 11PM the day before up to 1AM the day after. In figure \ref{savitzky_4} we increased this margin to 4 hours at each end. While the 1 extra hour does little to improve the smoothing and even leads to a worse result in one case, the effect disappears completely with the 4 additional hours. We use this 4 hour buffer for all smoothings. Furthermore we decided to use the larger window size, the smoother graphs make it easier on the eyes to compare multiple smoothings on one graph. Additionally effects such as the drop during lunchtime were often not seen with tighter window sizes.\\
\newpage %Required to fix wrapfigure formatting
\begin{wrapfigure}{r}{0.35\textwidth}
\centering
\captionsetup{justification=centering}
\includegraphics[width=0.25\textwidth]{legend.png}
\caption{Legend:\\ days of the week}
\label{legend}
\end{wrapfigure}
Using an adapted version of Simon W{\"u}nderlich's FFT$\_$eval code, Python's Matplotlib library, the mathematical formulas mentioned above and a Savitzky-Golay filter with window size 361 we graphically represent our data points per node. For one week (Monday 15 May - Sunday 21 May) we plot each day's smoothing, for easy comparison between the days. For the 2.4 GHz band we perform this separately for the non-overlapping channels 1, 6 and 11. Each color in the plot represents a day of the week, using the legend in figure \ref{legend}. All other plots use the same legend, although days may be missing.

@@

\subsection{5 GHz band}
\subsubsection{Recurring patterns between days}
\begin{figure}
\begin{minipage}{0.47\textwidth}
    \centering
	\includegraphics[width=\textwidth]{images/2_4_GHz/node1_2017-05-17_chan1_image}
    \caption{Node 1, ch. 1, 13/05 - 17/05} \label{node1-1}
\end{minipage}\hfill
\begin{minipage}{0.47\textwidth}
    \centering
	\includegraphics[width=\textwidth]{images/2_4_GHz/node1_2017-05-17_chan6_image}
    \caption{Node 1, ch. 6, 13/05 - 17/05} \label{node1-6}
\end{minipage}\hfill
\begin{minipage}{0.47\textwidth}
    \centering
	\includegraphics[width=\textwidth]{images/2_4_GHz/node1_2017-05-17_chan11_image}
    \caption{Node 1, ch. 11, 13/05 - 17/05} \label{node1-11}
\end{minipage}\hfill
\begin{minipage}{0.47\textwidth}
\centering
\includegraphics[width=\textwidth]{images/5_GHz/cot-node3-student_2017-05-22_chan36_image.pdf}
\caption{Node 3, ch. 36, 15/05 - 22/05}
\label{cot3_5ghz_pattern_chan36}
\end{minipage}\hfill
\end{figure}
The data gathered during the 5 GHz measurements proved to be less interesting than the data we gathered in the 2.4 GHz measurements. Most of the channels we measured turned out to be silent or contain very little signal. This is partly to be expected. Many common communication devices only use the 2.4 GHz frequency band and many routers in use do not yet support the 5 GHz band. Furthermore, any communication that does occur in this band can be more spread out as the band is a lot wider than the 2.4 GHz band. For the most part, the channels that did show traffic did not show any significant patterns or recurring phenomena. In most cases the received signal strength was very stable on all channels. We did notice some remarkable behaviour though. For example, node 3 measured a lower signal strength for some time around noon on both May 17\textsuperscript{th} and May 22\textsuperscript{nd} on channel 56, in figure \ref{cot3_5ghz_pattern_chan36}. The reason for this however is not as clear as with the daily peaks experienced in the 2.4 GHz It might be that the decrease occurs every couple of days, for example due to a router switching channels temporarily. It could also simply be an anomaly. For this node, channel 36 was the only somewhat active channel. All other channels saw barely any traffic in the 5 GHz band. \\\\
It is worth noting that, apart from channel 36, there is some difference in the stability of the received signal between the outdoor node and the indoor nodes. The outdoor node has at least some activity on every channel whereas most of the indoors nodes show the bare minimum of activity for most channels. An example of the outdoors node's activity on channel 56 versus an indoors node on that same channel is shown in figures \ref{node1_5ghz_traffic_chan56} and \ref{node9_5ghz_traffic_chan56}. There could be some outside source of slight intermittent interference. As we only had access to one outdoors node we cannot determine whether this generally occurs with outside nodes, or is specific to this one.

@@

There were some other interesting phenomena we observed in the 5 GHz channels. For example, node 6 shows a sudden drop in activity on both channels 56 and 64, coinciding with an increase in traffic on channels 112 and 120, as seen in figures \ref{node6_5ghz_traffic_chan56} up to \ref{node6_5ghz_traffic_chan120}. Looking at a random spectral scan from before and after the shift, we indeed see the same behaviour. There was clear activity in channels 56 and 64 before the shift, seen in figure \ref{cot6_50_early}.
\begin{figure}[ht]
\begin{minipage}{0.47\textwidth}
	\centering
	\includegraphics[width=\textwidth]{images/5_GHz/cot-node6-student_2017-05-20_chan56_image.pdf}
	\caption{Node 6, ch. 56, 14/05 - 20/05}
	\label{node6_5ghz_traffic_chan56}
\end{minipage}\hfill
\begin{minipage}{0.47\textwidth}
	\centering
	\includegraphics[width=\textwidth]{images/5_GHz/cot-node6-student_2017-05-20_chan64_image.pdf}
	\caption{Node 6, ch. 64, 14/05 - 20/05}
	\label{node6_5ghz_traffic_chan64}
\end{minipage}\hfill
\begin{minipage}{0.47\textwidth}
    \centering
    \includegraphics[width=\textwidth]{images/5_GHz/cot-node6-student_2017-05-20_chan112_image.pdf}
    \caption{Node 6, ch. 112, 14/05 - 20/05}
    \label{node6_5ghz_traffic_chan112}
\end{minipage}\hfill
\begin{minipage}{0.47\textwidth}
    \centering
    \includegraphics[width=\textwidth]{images/5_GHz/cot-node6-student_2017-05-20_chan120_image.pdf}
    \caption{Node 6, ch. 120, 14/05 - 20/05}
    \label{node6_5ghz_traffic_chan120}
\end{minipage}\hfill
\end{figure}
 While there is some very slight activity in channel 60, so few samples contained any trace of it that it was filtered out. Later on, figure \ref{cot6_50_late} shows that this traffic signature has moved to channels 112 and 120, now with the channel in between showing no traffic at all. We are not sure what device exactly causes this, but it is clear that it was at some point switched to another channel. This shows that the activity in a channel can change at a moment's notice, and indicates that systematic or even constant monitoring of the spectrum is not without merit. We do note that the received high signal is still only -70dBm, meaning there is still room for communication. Furthermore the spectral scan samples show that those channels (i.e. 112 and 120) are still silent for a large fraction of the time.

@@

\end{figure}\\
\subsection{Implications of high signal strength}
A high or unstable measured signal strength does not necessarily make a channel unusable for a City of Things node. First of all, we note that all received signal strengths previously reported was due to activity caused by external sources (i.e. not the node itself). When considering using the node for communication, all those signals are considered as noise. As long as this noise is lower than the signal strength used for the node's communication, there is no issue. While a constant noise of -20dBm will make communication very difficult, noise of -70dBm, -60dBm or even -50dBm should be no problem for close-range communication, as the node and its peer could probably generate a stronger signal. Furthermore it is important to note that in many cases the measured high signal strength was not constant. As it was probably caused by a communication device (likely Wi-Fi), the signal source remains silent for a significant fraction of the time. Furthermore such sources of potential interference probably implement some sort of listen-before-talk, further facilitating communication with the node in the presence of such sources. It is however difficult to predict exactly how well communication will work for a specific situation without actually trying it out.
\subsection{Validation}
While there is no official confirmation from the chip manufacturer that our interpretation of the spectral scan data is correct, we are confident that our interpretation gives at least a solid estimation of spectrum activity. First of all the experiment in section \ref{accuracy} shows that when we add a source of RF activity near the node, this is clearly seen in the spectral scan at exactly the right frequency. \\
Furthermore we discovered daily patterns where we expected them. The Middelheim offices are busy during the daytime on weekdays but fairly silent during the weekends and nights. While not shown in the plots above, holidays were just as quiet as weekends.
\section{Conclusions}
In this report we presented our findings on activity in the RF spectrum in the 2.4 GHz and 5 GHz bands in the city of Antwerp. One clear theme in our results is that the 5 GHz is overall quieter than the 2.4 GHz band. This is not surprising. The 5 GHz band is a lot wider than the 2.4 GHz band, and its signals do not carry as far. Furthermore, assuming that a lot of interference originated from Wi-Fi routers, one factor could be that many active routers and other devices do not yet support 5 GHz. In addition, technologies such as Bluetooth, which could also be creating signals, only affect 2.4 GHz.\\ \\
In some environments we notice a daily pattern where the signal strength is about 20dBm higher during the day compared to the night. As this was mainly observed on nodes in office spaces, this effect disappeared during the weekend. Other nodes show absolutely no difference between daytime and nighttime. Often this is when the spectrum is just very quiet. One interesting observation is that channels 6 and 11 were never busier than channel 1. \\
Furthermore, we found nothing indicating that City of Things network would experience considerable interference in the 5 GHz band, especially for the higher channel numbers. When operating on the 2.4 GHz band we observed more activity that could, in some places and on some channels, hinder communication with faraway hosts. However there was always at least one channel fairly silent and thus usable.
\bibliographystyle{IEEEtran}
\bibliography{IEEEabrv,./bibliography}
